 Casting multiple shadows: high-dimensional interactive data visualisation with tours and embeddings

### 2020-04-17 paper.Rmd
first draft of DR section

In the following section, we provide a brief overview of the goals
of DR methods with respect to the data analyst, in addition to their high level the 
mathematical details. To begin we suppose the data is in the form of an $n \times p$ 
rectangular matrix of real numbers, $X$, where $n$ is the number of  observations
in $p$ dimensions. The purpose of any DR algorithm is to find low-dimensonal
representation of the data, $Y$, such that $Y$ is an $n \times d$ matrix 
where $d \ll p$. The aspiration is that the DR procedure will remove noise
in the data while revealing hidden patterns and structure.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input 
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation. 
In this paper we restrict our attention to two methods that are commonly
used in the literature: t-distributed stochastic neighbor embedding (t-SNE) and
uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
the pairwise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kullback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similarities in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions.

- there's a lot of parameters to tweak
- emphasises locality, distance between clusters of points can be misleading
- size of clusters can be misleading
- may require a few different runs to capture topology of data
- if there are clusters it will find them

@@

We propose using tours as part of an analyst's workflow in in performing
dimensionality reduction tasks. We have made extensive use of ensemble
graphics, that is aligning related plots alongside each other to provide
context. As we will see in the case studies, this allows analysts to quickly
compare views from embedding methods and allows them to see how the embedding
method alters the global structure of their data. Using ensembles allows the
use of interaction techniques, that allow analysts to perform cluster
orientation tasks via linking multiple views. This approach allows our
interface, to achieve the three principles for interactive high-dimensional
data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
and making comparisons.

@@

We have combined the tour view in a side by side layout with a scatter plot
view that represents the output of an embedding algorithm.  These views are
linked; analysts can brush regions or highlight collections of points in either
view. Linked highlighting can be performed when points have been previously
labeled according to some discrete structure, i.e. cluster labels are
available. This is achieved via the analyst clicking on groups in the legend,
which causes unselected groupings to have their points become less opaque.
Consequently, simple linked highlighting can alleviate a known downfall of
methods such as UMAP or t-SNE: that is distances between clusters are
misleading. By highlighting corresponding clusters in the tour view, the
analyst can see the relationship between clusters, and therefore obtain a more
accurate representation of the topology of their data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether, points are near or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbors graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by ...
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively. We propose this can be
done visually using our side-by-side tour and embedding views. Some quality
checks, such as finding distortions can be identified using simple linked
brushing, however we can also interrogate them via spatial brushes and brush
composition.

@@

Using a linked neighbourhood brush, we can visually investigate the nearest
neighbour relationships in the high-dimensional space via brushing in the
embedding view. The user can select the number of nearest neighbours directly,
and modify the distance metric used for determining the neighbours. This allows
users to interrogate the stability of clusters generated in the embedding view.
Multiple brushes can be used to pose queries in either the tour view or the
embedding view; interface controls allow these brushes to combine using logical
operators such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple color highlights, allowing analysts to get a more holistic view of the
effect of an embedding algorithm.

>>>>

In the following section, we provide a brief overview of the goals
of DR methods with respect to the data analyst, in addition to their high level the 
mathematical details. To begin we suppose the data is in the form 
rectangular matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, 
where $n$ is the number of observations in $p$ dimensions. 
The purpose of any DR algorithm is to find low-dimensonal
representation of the data, $Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that
$Y$ is an $n \times d$ matrix where $d \ll p$. The hope of the analyst is that
the DR procedure to produces $Y$ will remove noise in the original dataset while
retaining latent structure.

@@

For non-linear methods $Y$ is generated via a pre-processed form of the input 
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation. 
Here we restrict our attention to two recent methods that are commonly
used in the literature: t-distributed stochastic neighbor embedding (t-SNE) and
uniform manifold alignment and projection (UMAP). The t-SNE algorithm estimates
the pairwise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pairwise distances between  
$\mathbf{x_i}$ and $\mathbf{x_j}$  to similiarties using a Gaussian kernel:

$$
p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 / 2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 / 2\sigma_i^2)}
$$

The conditional probabilities are then normalised and symmetrised to form a 
joint probability  distrubition via averaging:

$$
p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n}
$$

The variance parameter of the 
Gaussian kernel is controlled by the analyst using a fixed value of perplexity
for all observations:

$$
\text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i}))
$$
As the perplexity increases, $sigma_^2_{i}$ increases, until its bounded
above by the number of observations , $n-1$, in the data, corresponding to  
$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a 
nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all 
observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].

Next, the target low-dimensional space, $Y$, pairwise distances between 
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability 
distrubtion using a Cauchy kernel:

$$
q_{ij} = \frac{w_{ij}}{Z} \\
\text{where } w_{ij} = \frac{1}{ 1 + \lVert \mathbf{y_i - y_j} \rVert ^ 2} 
\text{ and } Z = \sum_{k \ne l} w_{kl}.
$$

The resulting configuration $Y$ is
the one that minimizes the Kullback-Leibler divergence between 
the probability distrubtions formed via similiarties of observations in $X$, $\mathcal{P}$ 
and similiarities of observations in $Y$, $\mathcal{Q}$:

$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log \frac{p_{ij}}{q_{ij}}$$

Re-writing the loss function in terms of attractive (right) and repulsive (left) 
forces we obtain:

$$
\mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} + \log\sum_{i \ne j} w_{ij}
$$

So when the the loss function is minimised this corresponds to large
attractive forces, that is, the pairwise distances in $Y$ are small when
there are non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The
repulsive force should also be small when the loss function is minimised, 
that is, when pairwise distances in $Y$ are large irregardless of the magnitude
of the corresponding distances in $X$.

Taken together, these details reveal the sheer number of decisions that an 
anlayst must make. How does one choose the perplexity? How should the parameters
that control the optimisation of the loss function (done with stochastic
gradient descent), like the number of iterations, or early exaggeration (
a multiplier of the attractive force at the beginning of the optimisation),
or the learning rate be selected?  It is a known
problem that t-SNE can have trouble recovering topology and that
configurations can be highly dependent on how the algorithm is initialised and
parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. If the goal is cluster
orientation a recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. However,
the cluster sizes and their relative orientation from a $t-SNE$ view can be 
misleading perceptually, due to the algorithms emphasis on locality.

Another recent method UMAP has seen a large rise in popularity (at least in
single cell transcriptomics) [@McInnes2018-co]. is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similarities in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. It appears to suffer from the same perceptual issues
as t-SNE, however it supposedly preservers global structure better
than t-SNE [@Coenen2019-to].

@@

We propose using tours as part of an analyst's workflow when using embedding
methods like t-SNE. We have made extensive use of ensemble
graphics, that is aligning related plots alongside each other to provide
context. As we will see in the case studies, this allows analysts to quickly
compare views from embedding methods and allows them to see how the embedding
method alters the global structure of their data. Using ensembles allows the
use of interaction techniques, that allow analysts to perform cluster
orientation tasks via linking multiple views. This approach allows our
interface, to achieve the three principles for interactive high-dimensional
data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
and making comparisons.

@@

We have combined the tour view in a side by side layout with a scatter plot
view that represents the output of an embedding algorithm.  These views are
linked; analysts can brush regions or highlight collections of points in either
view. Linked highlighting can be performed when points have been previously
labelled according to some discrete structure, i.e. cluster labels are
available. This is achieved via the analyst clicking on groups in the legend,
which causes unselected groupings to have their points become less opaque.
Consequently, simple linked highlighting can alleviate a known downfall of
methods such as UMAP or t-SNE: that is distances between clusters are
misleading. By highlighting corresponding clusters in the tour view, the
analyst can see the relationship between clusters, and therefore obtain a more
accurate representation of the topology of their data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether, points are near or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by ...
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively. We propose this can be
done visually using our side-by-side tour and embedding views. Some quality
checks, such as finding distortions can be identified using simple linked
brushing, however we can also interrogate them via spatial brushes and brush
composition.

@@

Using a linked neighbourhood brush, we can visually investigate the nearest
neighbour relationships in the high-dimensional space via brushing in the
embedding view. The user can select the number of nearest neighbours directly,
and modify the distance metric used for determining the neighbours. This allows
users to interrogate the stability of clusters generated in the embedding view.
Multiple brushes can be used to pose queries in either the tour view or the
embedding view; interface controls allow these brushes to combine using logical
operators such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple colour highlights, allowing analysts to get a more holistic view of the
effect of an embedding algorithm.
